The Role of Variability in Learning Generalization: A Computational Modeling Approach

### 2024-04-22 Sections/full.qmd
add docx version; fix igas figure text; re-render everything.

---
title: "The Role of Variability in Learning Generalization: A Computational Modeling Approach"
author:
- name: Thomas Gorman
  affiliation: Indiana University
page-layout: full
code-fold: true
code-tools: true
toc: true
toc-location: body
#toc-title: Contents
toc-depth: 4
cache: true
format-links: [hikmah-manuscript-pdf, gfm]
execute: 
  warning: false
  eval: true
format:
  html: default
  hugo-md:
    echo: false
    include: true
    html-math-method: mathjax
    output-file: diss_full_hugo.md
  gfm:
    echo: false
    output: false
    output-file: diss_full_gfm.md
  hikmah-manuscript-pdf:
    echo: false
    output-file: "tg_full_disser.pdf"
    mainfont: "Linux Libertine O"
    mainfontoptions:
      - "Numbers=Proportional"
      - "Numbers=OldStyle"
    mathfont: "Libertinus Math"
prefer-html: true
---

@@

```{r out.width = "60%", echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
#| label: fig-IGAS_Methods
#| fig.cap: The stimuli of the task consisted of a blue ball, which the participants would launch at the green target, while avoiding the red barrier. On each trial, the ball would appear in the center of the orange square, with the position of the orange square varying between experimental conditions. Participants were constrained to release the ball within the square.

@@

```{r}
#| label: fig-e2testa
#| fig.cap: Testing phase performance from each of the six testing positions. The six constant conditions are averaged together into a single constant group, compared against the single varied-trained group.B) Transfer performance from each of the 6 throwing locations from which all participants were tested. Each bar represents performance from one of seven distinct training groups (six constant groups in red, one varied group in blue). The x axis labels indicate the location(s) from which each group trained. Lower values along the y axis reflect better performance at the task (closer distance to target center). Error bars indicate standard error of the mean. 
#| fig-width: 11
#| fig-height: 9

@@

kable(tab2,escape=FALSE,booktabs=TRUE,col.names=c("Position","Constant","Varied"),align=c("l"))  %>% kableExtra::kable_styling(position="left") #%>% kable_classic() #%>% footnote(general=captionText,general_title = "")

@@

```{r}
#| label: fig-taskSpace
#| fig.cap: A) A visual representation of the combinations of throw parameters (x and y velocities applied to the ball at launch), which resulted in target hits during the testing phase. This empirical solution space was compiled from all of the participants in Experiment 2. B) shows the solution space within the context of all of the throws made throughout the testing phase of the experiment.
#| fig-width: 11
#| fig-height: 9
#| 
taskspace <- e2 %>% filter(AbsDistFromCenter<900)
taskspace$hitOrMiss <- ifelse(taskspace$trialType==11,"Hit Target","Missed Target")

@@

```{r}
#| label: fig-Toy-Model-dis
#| fig.cap: A simple model depicting the necessity of both of two separately fit generalization parameters, c, and a positive distance between training and testing contexts, in order for an instance model to predict a pattern of varied training from stimuli 400 and 800 outperforming constant training from position 600 at a test position of 600. For the top left panel, in which the generalization model assumes a single c value (-.008) for  both varied and constant conditions, and identical contexts across training and testing, the equation which generates the varied condition is - Amount of Generalization = $e^{(c\cdot|x-800|)} + e^{(c\cdot|x-400|)}$, whereas the constant group generalization is generated from $2\cdot e^{(c\cdot|x-600|)}$. For the top right panel, the c constants in the original equations are different for the 2 conditions, with $c=-.002$ for the varied condition, and $c=-.008$ for the constant condition. The bottom two panels are generated from identical equations to those immediately above, except for the addition of extra distance (100 units) to reflect the assumption of some change in context between training and testing conditions. Thus, the generalization model for the varied condition in the bottom-right panel is of the form - Amount of Generalization = $e^{(c_{varied}\cdot|x-800|)}+e^{(c_{varied}\cdot|x-400|)}$ .
#| fig.width: 11
#| fig.height: 11.5

@@

e1_tab %>%
  group_by(`Task Stage`, Fit_Method, Model, condit) %>%
  summarize(ME = mean(mean_error), .groups = "drop") %>%
  pivot_wider(
    names_from = c(Model, condit),
    values_from = ME,
    names_sep = "_"  # Add this line to specify the separator for column names
  ) %>%
  rename("Fit Method" = Fit_Method) %>%
  gt() %>%
  cols_move_to_start(columns = c(`Task Stage`)) %>%
  cols_label(
    `Task Stage` = "Task Stage"
  ) %>%
  fmt_number(
    columns = starts_with("ALM") | starts_with("EXAM"),
    decimals = 2
  ) %>%
  tab_spanner_delim(delim = "_") %>%
  tab_style(
    style = cell_fill(color = "white"),
     locations = cells_body(columns = everything(), rows = everything())
  ) %>%
  tab_style(
    style = cell_borders(sides = "top", color = "black", weight = px(1)),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.font.size = 12,
    heading.title.font.size = 14,
    heading.subtitle.font.size = 12,
    quarto.disable_processing = TRUE,
    column_labels.padding = 2,
    data_row.padding = 2
  )
```

@@

e23_tab <- rbind(e2_tab |> mutate(Exp="E2"), e3_tab |> mutate(Exp="E3")) 
gt_table <- e23_tab %>%
  pivot_wider(
    names_from = c(Exp, Model, condit),
    values_from = mean_error,
    names_glue = "{Exp}_{Model}_{condit}"
  ) %>%
  arrange(Fit_Method, `Task Stage`) %>%
  gt() %>%
  cols_move_to_start(columns = `Task Stage`) %>%
  cols_label(`Task Stage` = "Task Stage") %>%
  fmt_number(columns = matches("E2|E3"), decimals = 1) %>%
  tab_spanner_delim(delim = "_") %>%
  tab_style(
    style = list(
      cell_fill(color = "white"),
      cell_borders(sides = "top", color = "black", weight = px(1))
    ),
    locations = cells_body(columns = everything(), rows = everything())
  ) %>%
  tab_options(
    table.font.size = 12,
    heading.title.font.size = 14,
    heading.subtitle.font.size = 12,
    quarto.disable_processing = TRUE,
    column_labels.padding = 2,
    data_row.padding = 2
  )
gt_table
```

@@

```{r}
#| label: tbl-htw-ee-e23
#| tbl-cap: "Results of Bayesian Regression models predicting model error as a function of Model (ALM vs. EXAM), Condition (Constant vs. Varied), and the interaction between Model and Condition. The values represent the estimate coefficient for each term, with 95% credible intervals in brackets.  The intercept reflects the baseline of ALM and Constant. The other estimates indicate deviations from the baseline for the EXAM mode and varied condition. Lower values indicate better model fit."

@@

bct_table <- bct %>%
  mutate(
    across(c(Estimate, `95% CrI Lower`, `95% CrI Upper`), ~ round(., 2)),
    pd = round(pd, 2)
  ) %>%
  gt() %>%
  # tab_header(
  #   title = "Bayesian Model Results",
  #   subtitle = "Estimates and Credible Intervals for Each Term Across Experiments"
  # ) %>%
  cols_label(
    exp = "Experiment",
    Term = "Term",
    Estimate = "Estimate",
    `95% CrI Lower` = "95% CrI Lower",
    `95% CrI Upper` = "95% CrI Upper",
    pd = "pd"
  ) %>%
  fmt_number(
    columns = c(Estimate, `95% CrI Lower`, `95% CrI Upper`),
    decimals = 1
  ) %>%
  fmt_number(
    columns = pd,
    decimals = 2
  ) %>%
  tab_spanner(
    label = "Credible Interval",
    columns = c(`95% CrI Lower`, `95% CrI Upper`)
  ) %>%
  tab_style(
    style = list(
      #cell_fill(color = "lightgray"),
      cell_text(weight = "bold"), 
      cell_fill(color = "white"),
      cell_borders(sides = "top", color = "black", weight = px(1))
    ),
    locations = cells_body(
      columns = c(Estimate, pd),
      rows = Term=="ModelEXAM:conditVaried"
    )
  ) %>%
   tab_row_group(
    label = "Experiment 3",
    rows = exp == "Exp 3"
  ) %>%
  tab_row_group(
    label = "Experiment 2",
    rows = exp == "Exp 2"
  ) %>%
  tab_row_group(
    label = "Experiment 1",
    rows = exp == "Exp 1"
  ) %>%
  tab_options(
    table.font.size = 12,
    heading.title.font.size = 14,
    heading.subtitle.font.size = 12,
    quarto.disable_processing = TRUE,
    column_labels.padding = 2,
    data_row.padding = 2
    #row_group.background.color = "gray95"
  )
bct_table

>>>>

---
title: "The Role of Variability in Learning Generalization: A Computational Modeling Approach"
author:
- name: Thomas Gorman
  affiliation: Indiana University
page-layout: full
code-fold: true
code-tools: true
toc: true
toc-location: body
toc-depth: 4
cache: true
format-links: [hikmah-manuscript-pdf, gfm, docx]
execute: 
  warning: false
  eval: true
format:
  html: default
  hugo-md:
    echo: false
    include: true
    html-math-method: mathjax
    output-file: diss_full_hugo.md
  gfm:
    echo: false
    #output: false
    output-file: diss_full_gfm.md
  hikmah-manuscript-pdf:
    echo: false
    output-file: "tg_full_disser.pdf"
    mainfont: "Linux Libertine O"
    mainfontoptions:
      - "Numbers=Proportional"
      - "Numbers=OldStyle"
    mathfont: "Libertinus Math"
  docx:
    echo: false
    prefer-html: true
    output-file: "TG_Full_Dissert.docx"
    toc: true
prefer-html: true
---

@@

```{r out.width = "60%", echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
#| label: fig-IGAS_Methods
#| fig-cap: The stimuli of the task consisted of a blue ball, which the participants would launch at the green target, while avoiding the red barrier. On each trial, the ball would appear in the center of the orange square, with the position of the orange square varying between experimental conditions. Participants were constrained to release the ball within the square.

@@

```{r}
#| label: fig-e2testa
#| fig-cap: Testing phase performance from each of the six testing positions. The six constant conditions are averaged together into a single constant group, compared against the single varied-trained group.B) Transfer performance from each of the 6 throwing locations from which all participants were tested. Each bar represents performance from one of seven distinct training groups (six constant groups in red, one varied group in blue). The x axis labels indicate the location(s) from which each group trained. Lower values along the y axis reflect better performance at the task (closer distance to target center). Error bars indicate standard error of the mean. 
#| fig-width: 11
#| fig-height: 9

@@

kable(tab2,escape=FALSE,booktabs=TRUE,col.names=c("Position","Constant","Varied"),align=c("l"))  #%>% kableExtra::kable_styling(position="left") #%>% kable_classic() #%>% footnote(general=captionText,general_title = "")

@@

```{r}
#| label: fig-taskSpace
#| fig-cap: A) A visual representation of the combinations of throw parameters (x and y velocities applied to the ball at launch), which resulted in target hits during the testing phase. This empirical solution space was compiled from all of the participants in Experiment 2. B) shows the solution space within the context of all of the throws made throughout the testing phase of the experiment.
#| fig-width: 11
#| fig-height: 9
#| 
taskspace <- e2 %>% filter(AbsDistFromCenter<900)
taskspace$hitOrMiss <- ifelse(taskspace$trialType==11,"Hit Target","Missed Target")

@@

```{r}
#| label: fig-Toy-Model-dis
#| fig-cap: A simple model depicting the necessity of both of two separately fit generalization parameters, c, and a positive distance between training and testing contexts, in order for an instance model to predict a pattern of varied training from stimuli 400 and 800 outperforming constant training from position 600 at a test position of 600. For the top left panel, in which the generalization model assumes a single c value (-.008) for  both varied and constant conditions, and identical contexts across training and testing, the equation which generates the varied condition is - Amount of Generalization = $e^{(c\cdot|x-800|)} + e^{(c\cdot|x-400|)}$, whereas the constant group generalization is generated from $2\cdot e^{(c\cdot|x-600|)}$. For the top right panel, the c constants in the original equations are different for the 2 conditions, with $c=-.002$ for the varied condition, and $c=-.008$ for the constant condition. The bottom two panels are generated from identical equations to those immediately above, except for the addition of extra distance (100 units) to reflect the assumption of some change in context between training and testing conditions. Thus, the generalization model for the varied condition in the bottom-right panel is of the form - Amount of Generalization = $e^{(c_{varied}\cdot|x-800|)}+e^{(c_{varied}\cdot|x-400|)}$ .
#| fig.width: 11
#| fig.height: 11.5

@@

if (primary) {
e1_tab %>%
  group_by(`Task Stage`, Fit_Method, Model, condit) %>%
  summarize(ME = mean(mean_error), .groups = "drop") %>%
  pivot_wider(
    names_from = c(Model, condit),
    values_from = ME,
    names_sep = "_"  # Add this line to specify the separator for column names
  ) %>%
  rename("Fit Method" = Fit_Method) %>%
  gt() %>%
  cols_move_to_start(columns = c(`Task Stage`)) %>%
  cols_label(
    `Task Stage` = "Task Stage"
  ) %>%
  fmt_number(
    columns = starts_with("ALM") | starts_with("EXAM"),
    decimals = 2
  ) %>%
  tab_spanner_delim(delim = "_") %>%
  tab_style(
    style = cell_fill(color = "white"),
     locations = cells_body(columns = everything(), rows = everything())
  ) %>%
  tab_style(
    style = cell_borders(sides = "top", color = "black", weight = px(1)),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.font.size = 12,
    heading.title.font.size = 14,
    heading.subtitle.font.size = 12,
    quarto.disable_processing = TRUE,
    column_labels.padding = 2,
    data_row.padding = 2
  )
} else {

  e1_tab %>%
  group_by(`Task Stage`, Fit_Method, Model, condit) %>%
  summarize(ME = mean(mean_error), .groups = "drop") %>%
  pivot_wider(
    names_from = c(Model, condit),
    values_from = ME,
    names_sep = "_"  # Add this line to specify the separator for column names
  ) %>%
  rename("Fit Method" = Fit_Method) %>% kable(escape=F,booktabs=T)

  }
```

@@

e23_tab <- rbind(e2_tab |> mutate(Exp="E2"), e3_tab |> mutate(Exp="E3"))

if (primary) {
gt_table <- e23_tab %>%
  pivot_wider(
    names_from = c(Exp, Model, condit),
    values_from = mean_error,
    names_glue = "{Exp}_{Model}_{condit}"
  ) %>%
  arrange(Fit_Method, `Task Stage`) %>%
  gt() %>%
  cols_move_to_start(columns = `Task Stage`) %>%
  cols_label(`Task Stage` = "Task Stage") %>%
  fmt_number(columns = matches("E2|E3"), decimals = 1) %>%
  tab_spanner_delim(delim = "_") %>%
  tab_style(
    style = list(
      cell_fill(color = "white"),
      cell_borders(sides = "top", color = "black", weight = px(1))
    ),
    locations = cells_body(columns = everything(), rows = everything())
  ) %>%
  tab_options(
    table.font.size = 12,
    heading.title.font.size = 14,
    heading.subtitle.font.size = 12,
    quarto.disable_processing = TRUE,
    column_labels.padding = 2,
    data_row.padding = 2
  )
gt_table
} else {
  e23_tab %>%
  pivot_wider(
    names_from = c(Exp, Model, condit),
    values_from = mean_error,
    names_glue = "{Exp}_{Model}_{condit}"
  ) %>%
  arrange(Fit_Method, `Task Stage`) %>%
  kable(escape=F,booktabs=T)
}
```

@@

```{r}
#| label: tbl-htw-ee-e23
#| tbl-cap: "Results of Bayesian Regression models predicting model error as a function of Model (ALM vs. EXAM), Condition (Constant vs. Varied), and the interaction between Model and Condition. The values represent the estimate coefficient for each term, with 95% credible intervals in brackets.  The intercept reflects the baseline of ALM and Constant. The other estimates indicate deviations from the baseline for the EXAM mode and varied condition. Lower values indicate better model fit."
#| cache: false

@@

out_type <- knitr::opts_knit$get("rmarkdown.pandoc.to")
primary=out_type == "html" || out_type == "pdf" || out_type =="latex" || out_type == "docx"

if (primary) {
bct_table <- bct %>%
  mutate(
    across(c(Estimate, `95% CrI Lower`, `95% CrI Upper`), ~ round(., 2)),
    pd = round(pd, 2)
  ) %>%
  gt() %>%
  # tab_header(
  #   title = "Bayesian Model Results",
  #   subtitle = "Estimates and Credible Intervals for Each Term Across Experiments"
  # ) %>%
  cols_label(
    exp = "Experiment",
    Term = "Term",
    Estimate = "Estimate",
    `95% CrI Lower` = "95% CrI Lower",
    `95% CrI Upper` = "95% CrI Upper",
    pd = "pd"
  ) %>%
  fmt_number(
    columns = c(Estimate, `95% CrI Lower`, `95% CrI Upper`),
    decimals = 1
  ) %>%
  fmt_number(
    columns = pd,
    decimals = 2
  ) %>%
  tab_spanner(
    label = "Credible Interval",
    columns = c(`95% CrI Lower`, `95% CrI Upper`)
  ) %>%
  tab_style(
    style = list(
      #cell_fill(color = "lightgray"),
      cell_text(weight = "bold"), 
      cell_fill(color = "white"),
      cell_borders(sides = "top", color = "black", weight = px(1))
    ),
    locations = cells_body(
      columns = c(Estimate, pd),
      rows = Term=="ModelEXAM:conditVaried"
    )
  ) %>%
   tab_row_group(
    label = "Experiment 3",
    rows = exp == "Exp 3"
  ) %>%
  tab_row_group(
    label = "Experiment 2",
    rows = exp == "Exp 2"
  ) %>%
  tab_row_group(
    label = "Experiment 1",
    rows = exp == "Exp 1"
  ) %>%
  tab_options(
    table.font.size = 12,
    heading.title.font.size = 14,
    heading.subtitle.font.size = 12,
    quarto.disable_processing = TRUE,
    column_labels.padding = 2,
    data_row.padding = 2
    #row_group.background.color = "gray95"
  )
bct_table
} else {
  bct %>%
  mutate(
    across(c(Estimate, `95% CrI Lower`, `95% CrI Upper`), ~ round(., 2)),
    pd = round(pd, 2)
  ) %>% kable(booktabs = TRUE)

}
```
